\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2020-12-09 paper.tex
lights v3

\DeclareMathOperator{\argmin}{argmin}

@@

The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{AndBorGilKei-93}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.

@@

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for feature selection in the latent class membership as well as within the survival model, and sparse group lasso~\citep{simon2013sparse} for the fixed effect of each marker trajectories, allowing flexible representations of time.
Inference is achieved using an efficient Quasi-Newton Expectation Maximization algorithm.
The posterior estimates of the latent class membership probabilities are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters to adapt to the data complexity.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring.

@@

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
\begin{assumption}
Suppose that conditional on the random effects, each $Y_i^l$ is independent, and that the censoring times $T_i$ are independent of $b_i$.
\end{assumption}
This is a standard modeling assumption~\citep{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

Given~\eqref{eq:link-function}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g. in the Gaussian case) or the counting measure (e.g. in the logistic cases).
The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
\begin{equation}
	\label{eq:exp-family-density}
	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - \sum_{j=1}^{n_i} c_\phi(M_{ik, j}) + d_\phi(y_i) \big\},
\end{equation}
with 
\[ \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i} \] 
and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}. The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameters $\phi_l$, while parameters $\beta_k$ have to be estimated. Note that $d_\phi(\cdot)$ is related to the normalizing constant.
\begin{table}[htb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\toprule
 Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_{\phi_l}(\cdot)$ \\
 \midrule
 Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2\sigma_l^2$ \\ 
 Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
 Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto \log(z)$ & $z \mapsto e^z$ \\
 \bottomrule
\end{tabular}}
\caption[]{Examples of standard distributions that fit in the considered setting.}
\label{table:glm}
\end{table}

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
	\label{eq:ind-hyp}
	f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$. Hypothesis~\eqref{eq:ind-hyp} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
and
\begin{equation}
  \label{eq:b_density}
  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
\end{equation}

@@

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{en}, \eta} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vectors $\xi_k$ and $\gamma_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{v}_{\text{en}, \eta} = (1-\eta)\norm{v}_1 + \dfrac\eta2 \norm{v}_2^2 \]
for any vector $v$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{\beta_k^l}_2 + \tilde{\eta} \norm{\beta_k}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one. Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ (respectively $\gamma_k$) thus informs on the features involved in the $k$-th group membership (in the $k$-th group risk of event respectively).
The ridge part allows to handle potential correlation between time-independent features or between longitudinal outcomes.
Note that in practice, the intercept is not regularized.
On the other hand, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory with the right complexity.

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big) \\
  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - \sum_{j=1}^{n_i} c_\phi(M_{ik, j}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g(b_i, G_i) = \tilde g(bi)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(Gi)$ with $g_k : G_i \mapsto \ind{\{G_i=k\}}$.

In the case $g(b_i, G_i) = \tilde g(bi)$, one has
\begin{equation*}
  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1}  \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation*}
with
\begin{equation}
  \label{eq:Lambda_ik-def}
  \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \pi_{\xi_k^{(w)}}(x_i) \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-def} is intractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.

@@

\noindent The approximation of~\eqref{eq:Lambda_ik-def} is finally obtained by
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \pi_{\xi_k^{(w)}}(x_i) \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}

@@

Let us give some numerical details about the choices made for using the lights model in practice.

@@

% \appendix

% \begin{center}
% \LARGE \textbf{Appendices}
% \end{center}

% \section{Proofs}
% \label{sec:proofs}

>>>>

\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\Tr}{Tr}

@@

The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.

@@

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model, as well as for the fixed effect (allowing flexible representations of time). Indeed, this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model on one side. On the other hand, it yields sparsity for a given trajectory, namely feature selection.
Inference is achieved using an efficient and novel Quasi-Newton Monte Carlo Expectation Maximization algorithm.
% NOOOOON
% The posterior estimates of the latent class membership probabilities are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters to adapt to the data complexity.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring.

@@

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
\begin{assumption}
\label{indep-hyp-2}
Suppose that conditional on the random effects, each $Y_i^l$ is independent, and that the censoring times $T_i$ are independent of $b_i$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is a standard modeling assumption, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

Given~\eqref{eq:link-function}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g. in the Gaussian case) or the counting measure (e.g. in the logistic cases).
The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
\begin{equation}
	\label{eq:exp-family-density}
	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
\end{equation}
with 
\[ \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i} \] 
and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}. The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameters $\phi_l$, while parameters $\beta_k$ have to be estimated. Note that $d_\phi(\cdot)$ is related to the normalizing constant.
\begin{table}[htb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\toprule
 Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_\phi(\cdot)$ \\
 \midrule
 Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2 \sigma_l^2$ \\ 
 Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
 Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto \log(z)$ & $z \mapsto e^z$ \\
 \bottomrule
\end{tabular}}
\caption[]{Examples of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
\label{table:glm}
\end{table}

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
and
\begin{equation}
  \label{eq:b_density}
  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
\end{equation}

@@

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$ and $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{v}_{\text{en}, \eta} = (1-\eta)\norm{v}_1 + \dfrac\eta2 \norm{v}_2^2 \]
for any vector $v$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{v}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{v^l}_2 + \tilde{\eta} \norm{v}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
Note that in the $\gamma_k$ penalty, $l$ actually starts at 0 (and not 1), but for simplicity we do not change the notation. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independant features involved in the $k$-th group membership. The ridge part allows to handle potential correlation between features.
On the other hand, the sparse group lasso allows to perform feature selection for each trajectory : the support of $\gamma_k^l$ informs on the features involved in the $k$-th group risk of event for the $l$-th longitudinal outcome.
Finally, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory $l$ with the right complexity.
Note that in practice, the intercept is not regularized.

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big) \\
  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g(b_i, G_i) = \tilde g(bi)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:Lambda_ik-def}
  \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.

@@

\noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
\begin{equation}
  \label{eq:def-hat-E-g-tilde}
  \hat \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}

@@

Let us give some details about the choices made for using the lights model in practice.

@@

\appendix

\begin{center}
\LARGE \textbf{Appendices}
\end{center}

\section{Numerical details}
\label{sec:numerical-details}
